---
title: Lab 8
theme: cosmo
bibliography: references.bib
execute:
  freeze: auto  # re-render only when source changes
  eval: false
  echo: true
engine: julia
---

# Calculating introgression statistics - three ways.

Today's lab is all about looking for introgression. We'll use several approaches - first, Patterson's D statistics from lecture. Second, we'll use a tree based approaches (TWISST and TreeMix) to see if we can get corroborating results. But first, let's look at D-statistics.

## Patterson's D

This one is surprisingly simple to calculate by yourself given pretty much all of the functions you have already written. The below is a code example of how you can take allele frequencies for four populations, and from the calculate Patterson's D. The calculation is equivalent to ABBA-BABA, but using allele frequency differences. Suppose you have subset of the genotype matrix for each population, then: 

```{r}
#Calculate allele frequencies
p_1 = apply(pop_1_GT,1,freq_GT)
p_2 = apply(pop_2_GT,1,freq_GT)
p_3 = apply(pop_3_GT,1,freq_GT)
p_4 = apply(pop_4_GT,1,freq_GT)

#Find which alleles should be re-polarized based on outgroup
idx =  which(p_4 > 0.5)
#And repolarize
p1[idx] = 1-p1[idx]
p2[idx] = 1-p2[idx]
p3[idx] = 1-p3[idx]
p4[idx] = 1-p4[idx]

denom_sum = sum((p1-p2)*(p3-p4))
num_sum = sum((p1+p2-2*p1*p2)*(p3+p4-2*p3*p4))
D = denom_sum/num_sum
```

But, this won't return nice features like significance testing, or other bits that you'd want with your test. So, let's use a common software tool: [Dsuite](https://github.com/millanek/Dsuite).

Dsuite is great in its simplicity. It takes a `vcf` (we've got one!), and a table that lists which samples belong to which population. We'd continue working with the MUC19 data, but, problematically, we only have three populations and need four. So, we'll use the entirety of Chromosome 12 from the 1K genomes project, and use the opportunity to learn to manipulate vcf files.

First, you'll want to download the chr12 vcf. In a terminal, you can do this with:

```{bash}
wget https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/release/20130502/ALL.chr22.phase3_shapeit2_mvncall_integrated_v5b.20130502.genotypes.vcf.gz
bcftools index ALL.chr22.phase3_shapeit2_mvncall_integrated_v5b.20130502.genotypes.vcf.gz
```

This will download chr22 data (the smallest chromosome) for ~3606 samples, and will then index it. The index will be useful for us when we manipulate the file further. In the lab data folder, you'll find `samples_for_introgression.txt`, a file which contains the ~800 samples we'll look for introgression in. So, the first thing you'll want to do is to subset the vcf only to those samples.

Most vcf manipulations happen with `bcftools`. This is a suite of tools which allow you to filter/subset/query/analyze vcfs in a variety of ways. You will frequently use several of its capabilities when dealing with genomic data. In this case, we want to limit ourselves only to the samples in the populations we are interested in. We'll first need to make a list of these samples. You could import the table into R and output only the first column, or you can use `awk` to print only the first column as follows:

```{bash}
awk 'NR > 1{print $1}' samples_for_introgression.txt > samples_to_keep.txt
```

However you make your list of samples to keep, subsetting the vcf is quite easy:

```{bash}
bcftools view ALL.chr12.phase3_shapeit2_mvncall_integrated_v5b.20130502.genotypes.vcf.gz -S samples_to_keep.txt -Oz -o subsampled_chr22.vcf.gz
```

Now, many pieces of software will want you to have an index for the vcf, so index it as well:

```{bash}
bcftools index subsampled_chr22.vcf.gz
```

And now we have a file that is ready for Dsuite.


Dsuite also needs a file formatted as:

```{format}
Ind1_name   Pop1
Ind2_name   Pop2
...
Indx_name   Outgroup
```

In other words, we need to take the table from 1k genomes and do two things - return only columns 1 and 4, and then rename YRI samples to "Outgroup". Again, `awk` is super useful here, but we'll also use `sed`, another great command line tool for editing text. `awk` will output the first and fourth column, while sed will look for any place in the file that says "YRI" and replace it with "Outgroup". This could go wrong if any of the samples have "YRI" in the name, but thankfully that should not be the case.

```{bash}
awk 'NR > 1{print $1 "\t" $4}' samples_for_introgression.txt > dsuite_samples_table.tsv
sed -i 's/YRI/Outgroup/' dsuite_samples_table.tsv
```

Alright, now we are finally ready to run `Dsuite`, assuming you have installed it.

```{bash}
Dsuite Dtrios subsampled_chr22.vcf.gz dsuite_samples_table.tsv -o chr22_test
```

This will generate a couple of files, the one you will want is the one that ends in `_Dmin.txt`. This shows the arrangement of population triplets that produced the smallest possible signal of introgression. 

Here's a key for the populations:

ASW - American Southwest
ACB - AfroCarribean
CLM - Columbian
MXL - Mexican
FIN - Finnish

Remember that ideally, Patterson's D requires P1 and P2 to be sister. Is that true for all of the inferred arrangements? 

Your homework is to look at these results and _really_ think about them. Which of these introgression patterns make sense given human evolutionary history? Which are likely artifacts of data?

The other two methods for introgression will be presented _but you don't need to run them_. Frankly, they will take too long.


## Method 2: TWISST

STOP HERE! I JUST REALIZED THERE'S TWISST2, AND IT IS WAAAAY BETTER. BUT, I HAVEN'T FULLY TESTED IT. WILL HOPEFULLY HAVE BY LAB TIME.

[TWISST](https://github.com/simonhmartin/twisst) is not strictly a test for introgression, but it can be a useful tool to gauge whether (and where in the genome) different relationships between groups are supported. Twisst requires not just the vcf, but a series of gene trees across the genome. Constructing these is relatively straightforward once you start with a vcf, however. I've added a script to the `src` directory which will generate the commands to make gene trees in windows. It will need you to specify the chromosomes in your data and their lengths in a separate file (in this case, single line files).

```{bash}
echo "22" > chrs.txt
echo "51304566" > lengths.txt
bash genWind.sh subsampled_chr22.vcf.gz 10000 > cmds.txt
```

If you look at the head of the `cmds.txt` file, you'll see there's lots of individual blocks of code for each window:

```{output}
##22_0_wnd1000000
bcftools view -r 22:1-1000001 subsampled_chr12.vcf.gz > 22_0_wnd1000000.vcf
plink --vcf 22_0_wnd1000000.vcf --snps-only --allow-extra-chrs --export phylip-phased --out 22_0_wnd1000000
rm 22_0_wnd1000000.vcf
iqtree3 -redo -pre 22_0_wnd1000000 -nt AUTO -ntmax 4 -bb 1000 -m auto -s 22_0_wnd1000000.phy
```

There's 5173 windows here, so it will take it a while to actually run through them all in sequence, in principle. In practice - this data is a little funky and there's lots of the genome with, essentially, zero information, so many of the windows will just error out. Rather than make you generate these files yourself, I've added some trees to the lab data folder, just so you can see how TWISST works.

Just like `Dsuite`, twisst needs you to enumerate the groups. It has a slightly nicer format, in that you can have a file with group labels, and then seperately specify which groups you'd like to include in your testing. Here, we'll only look for introgression between ACB and ASW.

```{bash}
twisst.py -t trees.gz -w output_weights.csv -g ACB -g MXL -g ASW -g YRI --method complete --groupsFile dsuite_samples_table.tsv
```

The output will be a csv file which, for each tree, calculates how frequent different tree topologies are. Each column is a topology, and each row is a single tree. So, we can just plot the relative frequency of each of these as we run along the genome (in this case, a ~1Mb region on chr22). 

```{r}
twisst_out = read_csv("output_weights.csv")
# This is where I realized there's a new and better method, which outputs data quite differently.
```

## Treemix

Treemix is another approach to detecting introgression. It has many known issues, but it is quite popular. Rather than reinvent the wheel, I'll point you to the guide I used when I learned to use it, found [here](https://speciationgenomics.github.io/Treemix/). There's also many other useful little guides in that page, but beware many of the methods they suggest are either slightly or very out of date, and the software has updated. So don't be surprised if some elements don't work.